### 2013-05-18 draft.tex
fixed independent assessments section to be more subtle, more clear, 5% estimate

The use of wavelength-dispersive spectroscopy in combination with optical microscopy and EBSD allows characterization of the trace metal contents of specific calcite textures. By comparing the trace metal concentrations of primary textures to those of diagenetic textures, we can identify the trace metal signature of diagenetic alteration in our sample set. The same thin sections as above were coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer. Using the JEOL map analysis program, the beam was calibrated on crystals of dolomite, siderite, strontianite and rhodochrosite. Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nAmps, voltage of 15 kV, and a counting time of 80 milliseconds per pixel. Quantitative spot measurements were made using the JEOL quantitiative analysis program.

@@

Although we attempted to sample only well-preserved skeletal material for the carbonate clumped isotope measurements, the large mass of material required makes it very difficult to avoid including small amounts of other phases. The trace metal concentrations of bulk samples can be used as an indicator of the amount of sediment and recrystallized material included in the sample \citep{Azmy1998,Brand1980,Brand2012,Came2007,Finnegan2011,Shields2003}. To measure bulk trace metal concentrations, between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$). Standards were prepared from standard solutions of Sr, Mn and Fe with the same concentration of Ca and HNO$_3$ as the sample solutions to minimize matrix effects. Solutions were sampled using a Cetac ASX 260 autosampler, aspirated into the Ar plasma using a peristaltic pump, and then analyzed on a Thermo Scientific iCAP 6000 series inductively-coupled plasma optical emission spectrometer (ICP-OES). Between each set of ten unknown samples, three standard solutions (10 ppb, 100 ppb, and 1 ppm) were measured. Reproducibility was on average 3\% for Fe, 2\% for Mn and 3\% for Sr.

@@

The \deltac C values of fossil material, micrite and cement are highly consistent within beds (Figure \ref{section}). We resolve a 3$\permil$ positive excursion in \deltac C, beginning just below the boundary between the Lower and Upper Visby Beds, and continuing for at least 50 meters. The sample set does not capture the relaxation of this excursion, but our samples from the Slite Formation, beginning at 76 meters, record a return to lower \deltac C$_{\text{PDB}}$ values of approximately -0.5$\permil$. The \deltac C trend through the section is comparable to the brachiopod \deltac C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.

@@

The large scatter in temperatures also suggests variable preservation quality among the samples. Even brachiopods, which are commonly thought to be resistant to diagenetic alteration, have a range in temperature of more than 15\degrees C. All of the fossil samples have fairly high values of \deltao O, making alteration by meteoric water unlikely. This leads to a hypothesis that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at high temperature. We tested this hypothesis using independent methods to assess the preservation quality of representative samples.

@@

The concentrations of trace metals in fossil calcite have been widely used as indicators of recrystallization and contamination from cements and sediments. Diagenetic recrystallization of calcite is thought to enrich iron (Fe) and manganese (Mn) and deplete strontium (Sr) \citep{Brand1980,Shields2003}, so we measured the concentrations of these three elements in a large subset of our samples. The Mn and Sr concentrations of almost all of the brachiopods analyzed in this study fall within the most strictly defined limits for the Mn and Sr concentrations of pristine brachiopods (Figure \ref{Brand}). The rugose corals tend to have more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while the cements and micrites tend to have the most highly altered trace metal signatures.

The implications of the brachiopod trace metal compositions are inconsistent with the clumped isotope results. The brachiopods in this study have a range of temperatures of more than 15\degrees C, indicating variable preservation, but they would all be judged pristine or nearly pristine based on their trace metal compositions. We hesitate to extend this interpretation to rugose corals, because their trace metal composition is less well-studied than that of brachiopods. Regardless, the results in Figure \ref{Brand} imply that the carbonate clumped isotope thermometer is a more sensitive indicator of diagenetic alteration than are bulk trace metal compositions.

We further tested the sensitivity of trace metal composition to diagenetic alteration by investigating possible relationships between trace metal composition and temperature (Figure \ref{metals}). We tested brachiopods and rugose corals separately and combined with micrites and cements for the robustness of a linear correlation between trace element composition and temperature. We also tested for a linear correlation between the log of trace element composition and temperature. We performed a hypothesis test on each pair of variables to test the probability (p) that the observed correlation is due to random variation when the true correlation is zero. Only one pair of variables, the Sr concentration and temperature of rugose corals, has a p value of less than 0.05, meaning we can be 95\% confident that the correlation is real.

There are several possible explanations for the scarcity of strong correlations between trace metal concentrations and temperature. Our sample set may not include a sufficient number of highly altered samples to capture the full range in severity of alteration. Indeed, the brachiopods in this study have a much smaller range in trace metal concentrations than the brachiopods analyzed by \cite{Brand2012} (Figure \ref{Brand}). The variable temperatures of cements, depending on whether the cements formed early during shallow burial or later in the presence of high-temperature fluids, may also contribute to the lack of correlation between trace metal concentrations and temperatures.

An additional explanation for the lack of strong correlations between temperature and trace metal concentrations is provided by electron microprobe (E-probe analysis) of the trace metal signatures of recrystallization. A recrystallized region of a rugose coral septum is clearly visible as a dark area in reflected light (Figure \ref{EBSD_coral}A) and an area of aligned calcite crystals in an electron backscatter diffraction (EBSD) image (Figure \ref{EBSD_coral}B). However, the Fe and Mn concentrations of the recrystallized region are indistinguishable from those of the pristine coral skeleton (Figure \ref{EBSD_coral}C \& D). This is only one example, but it suggests that variations in trace metal concentrations may be dominated by contamination from cements and sediments, rather than recrystallization. If recrystallization is an important contributor to the elevated temperatures of fossil material, trace metals may not be expected to be reliable indicators of diagenetic alteration.

@@

In contrast to rugose corals, the EBSD image of a low-temperature brachiopod thin section demonstrates excellent preservation of original fibrous calcite textures (Figure \ref{EBSD_brach}A). Within the brachiopod shell, calcite crystals are elongated approximately parallel to the surface of the shell, as expected for the fibrous secondary layer. This preferential orientation is represented by the vertical band in the stereographic projection of crystallographic planes (Figure \ref{EBSD_brach}B). No regions of recrystallization are visible within the shell. There are large interlocking cement crystals in the upper left hand corner of the EBSD image, but these formed in the void space between the two valves, and would not have been included in the sample for clumped isotope analysis.

@@

The micro-scale studies of calcite textures in Figures \ref{EBSD_coral}--\ref{flakes} reveal relatively well-preserved samples, but do not rule out the inclusion of small amounts of late-phase calcite. Variable fractions of late-phase calcite likely contributed to the range in temperature among fossil samples. We only performed SEM and EBSD analysis on a small subset of samples, so we will not attempt to use this evidence to compare the relative preservation qualtiy of low and high-temperature samples. An important conclusion that can be drawn from our micro-scale studies is that the low temperature brachiopods imaged in Figures \ref{EBSD_brach} and \ref{flakes} probably contain only trace amounts of late-phase calcite. We therefore look to these samples as the most reliable records of ocean temperature and \deltao O during the Silurian.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_Brand-01.jpg}
\caption{Bulk manganese and strontium concentrations of the brachiopods, micrites and cements in this study. The brachiopods in this study have more strontium and less manganese than most of the brachiopods analyzed by Brand et al. (2012), which is also shown in this figure. Lines represent estimates by different authors of the range of pristine concentrations.}
\label{Brand}
\end{figure}

>>>>

The use of wavelength-dispersive spectroscopy in combination with optical microscopy and EBSD allows characterization of the trace metal contents of specific calcite textures. By comparing the trace metal concentrations of primary textures to those of diagenetic textures, we can identify the trace metal signature of diagenetic alteration in our sample set. The same thin sections as above were coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer (E-probe). Using the JEOL map analysis program, the beam was calibrated on crystals of dolomite, siderite, strontianite and rhodochrosite. Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nAmps, voltage of 15 kV, and a counting time of 80 milliseconds per pixel. Quantitative spot measurements were made using the JEOL quantitiative analysis program.

@@

Although we attempted to sample only well-preserved skeletal material for the carbonate clumped isotope measurements, the large mass of material required makes it very difficult to avoid including small amounts of other phases. The trace metal concentrations of bulk samples can be used as an indicator of the amount of sediment and recrystallized material included in the sample \citep{Azmy1998,Brand2012,Came2007,Finnegan2011}. Diagenetic recrystallization of calcite is thought to enrich iron (Fe) and manganese (Mn) and deplete strontium (Sr) \citep{Brand1980,Shields2003}, so we measured the bulk concentrations of these three elements in a large and representative subset of our samples. Between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$). Standards were prepared from standard solutions of Sr, Mn and Fe with the same concentration of Ca and HNO$_3$ as the sample solutions to minimize matrix effects. Solutions were sampled using a Cetac ASX 260 autosampler, aspirated into the Ar plasma using a peristaltic pump, and then analyzed on a Thermo Scientific iCAP 6000 series inductively-coupled plasma optical emission spectrometer (ICP-OES). Between each set of ten unknown samples, three standard solutions (10 ppb, 100 ppb, and 1 ppm) were measured. Reproducibility was on average 3\% for Fe, 2\% for Mn and 3\% for Sr.

@@

The \deltac C values of fossil material, micrite and cement are highly consistent within beds (Figure \ref{section}). We resolve a 3$\permil$ positive excursion in \deltac C, beginning just below the boundary between the Lower and Upper Visby Beds, and continuing for at least 50 meters. The sample set does not capture the relaxation of this excursion, but samples from the Slite Formation, beginning at 76 meters, record a return to lower \deltac C$_{\text{PDB}}$ values of approximately -0.5$\permil$. The \deltac C trend through the section is comparable to the brachiopod \deltac C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.

@@

The large scatter in temperatures suggests variable preservation quality among the samples. Even brachiopods, which are commonly thought to be resistant to diagenetic alteration, have a range in temperature of more than 15\degrees C. All of the fossil samples have fairly high values of \deltao O, making alteration by meteoric water unlikely. This leads to a hypothesis that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at high temperature. We tested this hypothesis using independent methods to assess the preservation quality of representative samples.

@@

Bulk trace metal measurements, which have been widely used as indicators of diagenetic alteration, suggest that all the brachiopods in this study are pristine. The bulk Mn and Sr concentrations of all brachiopods fall within previously defined limits for the Mn and Sr concentrations of pristine brachiopods (Figure \ref{Brand}). However, micro-scale studies of calcite fabrics, as well as clumped isotope measurements, demonstrate that there is a range in preservation quality among the brachiopods that cannot be identified on the basis of trace metal measurements alone. We first explore possible reasons for the insensitivity of trace metal concentrations to preservation quality. Then we consider EBSD and SEM evidence for the inclusion of diagenetic material in the typical fossil sample. Finally, we explore the use of clumped isotopes to distinguish the best-preserved samples from among those that would ordinarily all be identified as pristine based on their trace metal contents.

There are several possible explanations for the relative insensitivity of trace metal concentrations to diagenetic alteration. Our sample set may not include a sufficient number of highly altered samples to capture the full range in severity of alteration. This hypothesis is supported by a conspicuous lack of statistically robust correlations between trace metal composition and temperature (Figure \ref{metals}). We tested brachiopods and rugose corals separately and combined with micrites and cements for the robustness of a linear correlation between trace metal concentrations and temperature, as well as the log of trace metal concentrations and temperature. We performed a hypothesis test on each pair of variables to test the probability (p) that the observed correlation is due to random variation when the true correlation is zero. Only one pair of variables, the Sr concentration and temperature of rugose corals, has a p value of less than 0.05, meaning we can be 95\% confident that the correlation is real. While these tests suggest that the scatter in trace metal concentrations is a large cause of their insensitivity to diagenetic alteration, there is also the possibility that early void-filling cements, which would have cool temperatures but altered trace metal concentrations, interfere with the signal.

E-probe analysis suggests a more compelling reason for the insensitivity of trace metal concentrations: diagenetic recrystallization may not actually have any effect on Fe and Mn concentrations. A recrystallized region of a rugose coral septum is clearly visible as a dark area in reflected light (Figure \ref{EBSD_coral}A) and an area of aligned calcite crystals in an electron backscatter diffraction (EBSD) image (Figure \ref{EBSD_coral}B). However, the Fe and Mn concentrations of the recrystallized region are indistinguishable from those of the pristine coral skeleton (Figure \ref{EBSD_coral}C \& D). This is only one example, but it suggests that recrystallization may not affect trace metal concentrations, and instead, variations in trace metal concentrations may be dominated by contamination from cements and sediments. If recrystallization is an important contributor to the elevated temperatures of fossil material, trace metals may not be expected to be reliable indicators of diagenetic alteration.

@@

In contrast to rugose corals, the EBSD image of a low-temperature brachiopod thin section demonstrates excellent preservation of original fibrous calcite textures (Figure \ref{EBSD_brach}A). Within the brachiopod shell, calcite crystals are elongated approximately parallel to the surface of the shell, as expected for the fibrous secondary layer. This preferential orientation is represented by the vertical band in the stereographic projection of crystallographic planes (Figure \ref{EBSD_brach}B). No regions of recrystallization are visible within the shell. There are large interlocking cement crystals in the upper left hand corner of the EBSD image, but these formed in the void space between the two valves, and would not have been included in the sample for clumped isotope analysis. Based on visual inspection of the EBSD image of this brachiopod's secondary fibrous layer, we suggest that this layer contains a maximum of 5\% recrystallized material.

@@

The micro-scale studies of calcite textures in Figures \ref{EBSD_coral}--\ref{flakes} reveal relatively well-preserved samples, but do not rule out the inclusion of small amounts of late-phase calcite. Variable fractions of late-phase calcite likely contribute to the range in temperature among fossil samples. We only performed SEM and EBSD analysis on a small subset of samples, so we will not attempt to use this evidence to compare the relative preservation quality of low and high-temperature samples. An important conclusion that can be drawn from our micro-scale studies is that the low temperature brachiopods imaged in Figures \ref{EBSD_brach} and \ref{flakes} probably contain only trace amounts of late-phase calcite, likely less than 5\% of the sample. We therefore look to these samples as the most reliable records of ocean temperature and \deltao O during the Silurian.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_Brand-01.jpg}
\caption{Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study. The brachiopods in this study have less Mn and more Sr than most of the brachiopods analyzed by Brand et al. (2012), which are also shown in this figure. Lines represent estimates by different authors of the range of pristine concentrations defined for brachiopods. Rugose corals generally have more Fe (not shown), more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while cements and micrites tend to have the most highly altered trace metal signatures.}
\label{Brand}
\end{figure}
